\textbf{Unshuffling Permutations}

### 2016-04-05 slides/slides.tex
pictures

@@

  \begin{block}{Remarks}
    \begin{itemize}
      \item
      We reduce from the problem of deciding wether a pattern occurs in a
      permutation.

@@

      \item
    \end{itemize}
  \end{block}
\end{frame}

@@

  \centering
  \begin{tikzpicture}[
    scale=.6,
    >=stealth',
    shorten >=1pt,
    main node/.style={align=center},
    cell/.style={draw,ultra thick,fill=black!5},
    structure link/.style={line width=1.5pt},
    pattern link/.style={line width=1.5pt,black!20},
    monotone/.style={draw,ultra thick,fill=black!5}
    ]
    % arcs
    % pi 1 - pi 2
    \draw [pattern link,->,>=latex'] (8.5,6) .. controls +(0,1) and +(0,2) .. (10.4,5);
    % pi 2 - sigma 2
    \draw [pattern link,->,>=latex'] (10.6,5) .. controls +(0,1) and +(0,3) .. (11.5,4);
    % 1
    \draw [structure link,->,>=latex'] (0.5,11) .. controls +(0,2.25) and +(0,1.25) .. (2.5,12);
  %       \node (claim edge 1) at (1.5,13.5) {Claim~\ref{claim:(nu_1, nu'_1)-edge}};
  %       \node (claim edge 1) at (1.5,13) {Claim~\ref{claim:all (nu_1, nu'_1)-edges}};
    % 2
    \draw [structure link,->,>=latex'] (1.5,2) .. controls +(0,1) and +(0,3) .. (6.5,1);
  %       \node (claim edge 2) at (4,4) {Claim~\ref{claim:(nu_2, nu'_2)-edge}};
  %       \node (claim edge 2) at (4,3.5) {Claim~\ref{claim:all (nu_2, nu'_2)-edges}};
    % 3
    \draw [structure link,->,>=latex'] (3.5,9) .. controls +(0,3) and +(0,1.5) .. (7.5,10);
  %       \node (claim edge 1) at (5.5,11.5) {Claim~\ref{claim:(nu_3, nu'_3)-edge}};
    % 4
    \draw [structure link,->,>=latex'] (5.5,8) .. controls +(0,1) and +(0,7) .. (9.5,3);
  %       \node (claim edge 1) at (8.5,8.5) {Claim~\ref{claim:(nu_4, nu'_4)-edge}};
    % sigma 1 - pi 1
    \draw [structure link,->,>=latex',red] (4.5,6) .. controls +(0,-3) and +(0,-2) .. (8.5,5);
    % \node (claim edge 1) at (6.5,4) {Claim~\ref{}};
    % nodes
    % 1
    \draw [cell] (0,10) -- (1,10) -- (1,11) -- (0,11) -- cycle;
    \draw [monotone] (0.5,10.5) node {\small  $\nu_1$};
    %\draw [draw,ultra thick,fill=black!5] (0,10) -- (1,10) -- (1,11) -- (0,11) -- cycle;
    \draw [cell] (2,11) -- (3,11) -- (3,12) -- (2,12) -- cycle;
    \draw [monotone] (2.5,11.5) node {\small  $\nu'_1$};
    % 2
    \draw [cell] (1,1) -- (2,1) -- (2,2) -- (1,2) -- cycle;
    \draw [monotone] (1.5,1.5) node {\small  $\nu_2$};
    \draw [cell] (6,0) -- (7,0) -- (7,1) -- (6,1) -- cycle;
    \draw [monotone] (6.5,0.5) node {\small  $\nu'_2$};
    % 3
    \draw [cell] (3,8) -- (4,8) -- (4,9) -- (3,9) -- cycle;
    \draw [monotone] (3.5,8.5) node {\small  $\nu_3$};
    \draw [cell] (7,9) -- (8,9) -- (8,10) -- (7,10) -- cycle;
    \draw [monotone] (7.5,9.5) node {\small  $\nu'_3$};
    % 4
    \draw [cell] (5,7) -- (6,7) -- (6,8) -- (5,8) -- cycle;
    \draw [monotone] (5.5,7.5) node {\small $\nu_4$};
    \draw [cell] (9,2) -- (10,2) -- (10,3) -- (9,3) -- cycle;
    \draw [monotone] (9.5,2.5) node {\small  $\nu'_4$};
    % sigma 1
    \draw [cell] (4,6) -- (5,6) -- (5,7) -- (4,7) -- cycle;
    \node [main node] (sigma1) at (4.5,6.5) {\small $\sigma'$};
    % pi 1
    \draw [cell] (8,5) -- (9,5) -- (9,6) -- (8,6) -- cycle;
    \node [main node] (pi1) at (8.5,5.5) {\small $\pi'$};
    % pi 2
    \draw [cell] (10,4) -- (11,4) -- (11,5) -- (10,5) -- cycle;
    \node [main node] (sigma2) at (10.5,4.5) {\small $\pi''$};
    % sigma 2
    \draw [cell] (11,3) -- (12,3) -- (12,4) -- (11,4) -- cycle;
    \node [main node] (pi2) at (11.5,3.5) {\small $\sigma''$};
    % grid
    \draw[step=1cm,black!30,ultra thin,fill=black!10] (0,0) grid (12,12);
    % row size
    \foreach \y/\N in {0.5/N_2,1.5/N_2,2.5/N_4,3.5/k,4.5/n,5.5/n+2,6.5/k+2,7.5/N_4,8.5/N_3,9.5/N_3,10.5/N_1,11.4/N_1} {
        \node at (-0.5,\y) (R\y) {\tiny $\N$};
    }
    % column size
    \foreach \x/\N in
    {0.5/N_1,1.5/N_2,2.5/N_1,3.5/N_3,4.5/k+2,5.5/N_4,6.5/N_2,7.5/N_3,8.5/n+2,9.5/N_4,10.5/n,11.5/k} {
        \node at (\x,-0.5) (C\x) {\tiny $\N$};
    }
  \end{tikzpicture}
\end{frame}

>>>>

\begin{frame}
  \frametitle{Permutation pattern}

  \begin{definition}
    A permutation $\pi$ is said to \emph{contain} the permutation $\sigma$ if there
    exists a subsequence of (not necessarily consecutive) entries of $\pi$ that
    has the same relative order as $\sigma$, and in this case $\sigma$ is said
    to be a pattern of $\pi$, written $\sigma \leq \pi$.

    Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$.
  \end{definition}

  \begin{exampleblock}{Example}
    \begin{itemize}
      \item
        The permutation $\pi = 391867452$ contains the pattern $\sigma = 51342$,
        as can be seen in the highlighted subsequence of
        $\pi = 3\mathbf{91}8\mathbf{674}52$
        (or $\pi = 3\mathbf{91}8\mathbf{67}4\mathbf{5}2$ or
        $\pi = 3\mathbf{91}8\mathbf{67}45\mathbf{2}$ or
        $\pi = 3\mathbf{91}867\mathbf{452}$.)

      \item
       Since the permutation $\pi = 391867452$ contains no increasing subsequence
       of length four, $\pi$ avoids $1234$.
    \end{itemize}
  \end{exampleblock}

\end{frame}

\begin{frame}
  \frametitle{Square permutation}

  \begin{definition}
    A permutation is said to be a \emph{square} if it can be obtained by
    shuffling two order-isomorphic patterns
  \end{definition}

  \begin{exampleblock}{Example}
    \begin{itemize}
      \item
        The permutation $\pi = 25167834$ is a square since it can obtained by
        shuffling two order-isomorphic (\emph{i.e.} $1342$)
        patterns $1683$ and $2574$ as can be seen in the highlighted subsequence of
        $\pi = \mathbf{25}16\mathbf{7}83\mathbf{4}$.

      \item
       $\pi = 4123$ and $1432$ are not squares.
    \end{itemize}
  \end{exampleblock}

\end{frame}

\begin{frame}
  \frametitle{Words}

  \begin{definition}
    The shuffle operator on words is defined inductively as follows:
    \begin{align*}
      u \shuffle \epsilon &= \epsilon \shuffle u = \{u\}
      \\
      (ua \shuffle vb) &= (ua \shuffle v)b \cup (u \shuffle vb)a
    \end{align*}
  \end{definition}

  \begin{block}{Key results}
    \begin{itemize}
      \item
      Given two words $v_1$ and $v_2$, the set $v_1 \shuffle v_2$
      can be computed in
      $O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)$
      time.

      \item
      Given three words $u$, $v_1$, and $v_2$, deciding if $u$ is a shuffle
      of $v_1$ and $v_2$ can be done in
      $O\left(n^2 / \log(n)\right)$ time.

      \item
      Given a word $u$, deciding if there is a word $v$ such that $u$ is
      in $v \shuffle v$ is \NPC.
    \end{itemize}
  \end{block}

\end{frame}

@@

  \begin{block}{Remarks}
    \begin{itemize}
      \item
      We reduce from the problem of deciding wether a pattern $\sigma$ occurs in a
      permutation $\pi$.

@@

      \item
      Construct a new permutation $\mu$ such that
      $\sigma$ occurs in $\pi$ if and only if
      there exists an oriented perfect matching $\mathcal{M}$ on $\mu$
      satisfying $\mathbf{P_1}$ and $\mathbf{P_2}$.
    \end{itemize}
  \end{block}
\end{frame}

@@

  \centering
  \begin{tikzpicture}[
    scale=.6,
    >=stealth',
    shorten >=1pt,
    main node/.style={align=center},
    cell/.style={draw,ultra thick,fill=black!5},
    structure link/.style={line width=1.5pt},
    pattern link/.style={line width=1.5pt,black!20},
    monotone/.style={draw,ultra thick,fill=black!5}
    ]
    % arcs
    % pi 1 - pi 2
    \draw [pattern link,->,>=latex'] (8.5,6) .. controls +(0,1) and +(0,2) .. (10.4,5);
    % pi 2 - sigma 2
    \draw [pattern link,->,>=latex'] (10.6,5) .. controls +(0,1) and +(0,3) .. (11.5,4);
    % 1
    \draw [structure link,->,>=latex'] (0.5,11) .. controls +(0,2.25) and +(0,1.25) .. (2.5,12);
    % 2
    \draw [structure link,->,>=latex'] (1.5,2) .. controls +(0,1) and +(0,3) .. (6.5,1);
    % 3
    \draw [structure link,->,>=latex'] (3.5,9) .. controls +(0,3) and +(0,1.5) .. (7.5,10);
    % 4
    \draw [structure link,->,>=latex'] (5.5,8) .. controls +(0,1) and +(0,7) .. (9.5,3);
    % sigma 1 - pi 1
    \draw [structure link,->,>=latex',red] (4.5,6) .. controls +(0,-3) and +(0,-2) .. (8.5,5);
    % nodes
    % 1
    \draw [cell] (0,10) -- (1,10) -- (1,11) -- (0,11) -- cycle;
    \draw [monotone] (0.5,10.5) node {\small  $\nu_1$};
    \draw [cell] (2,11) -- (3,11) -- (3,12) -- (2,12) -- cycle;
    \draw [monotone] (2.5,11.5) node {\small  $\nu'_1$};
    % 2
    \draw [cell] (1,1) -- (2,1) -- (2,2) -- (1,2) -- cycle;
    \draw [monotone] (1.5,1.5) node {\small  $\nu_2$};
    \draw [cell] (6,0) -- (7,0) -- (7,1) -- (6,1) -- cycle;
    \draw [monotone] (6.5,0.5) node {\small  $\nu'_2$};
    % 3
    \draw [cell] (3,8) -- (4,8) -- (4,9) -- (3,9) -- cycle;
    \draw [monotone] (3.5,8.5) node {\small  $\nu_3$};
    \draw [cell] (7,9) -- (8,9) -- (8,10) -- (7,10) -- cycle;
    \draw [monotone] (7.5,9.5) node {\small  $\nu'_3$};
    % 4
    \draw [cell] (5,7) -- (6,7) -- (6,8) -- (5,8) -- cycle;
    \draw [monotone] (5.5,7.5) node {\small $\nu_4$};
    \draw [cell] (9,2) -- (10,2) -- (10,3) -- (9,3) -- cycle;
    \draw [monotone] (9.5,2.5) node {\small  $\nu'_4$};
    % sigma 1
    \draw [cell] (4,6) -- (5,6) -- (5,7) -- (4,7) -- cycle;
    \node [main node] (sigma1) at (4.5,6.5) {\small $\sigma'$};
    % pi 1
    \draw [cell] (8,5) -- (9,5) -- (9,6) -- (8,6) -- cycle;
    \node [main node] (pi1) at (8.5,5.5) {\small $\pi'$};
    % pi 2
    \draw [cell] (10,4) -- (11,4) -- (11,5) -- (10,5) -- cycle;
    \node [main node] (sigma2) at (10.5,4.5) {\small $\pi''$};
    % sigma 2
    \draw [cell] (11,3) -- (12,3) -- (12,4) -- (11,4) -- cycle;
    \node [main node] (pi2) at (11.5,3.5) {\small $\sigma''$};
    % grid
    \draw[step=1cm,black!30,ultra thin,fill=black!10] (0,0) grid (12,12);
    % row size
    \foreach \y/\N in {0.5/N_2,1.5/N_2,2.5/N_4,3.5/k,4.5/n,5.5/n+2,6.5/k+2,7.5/N_4,8.5/N_3,9.5/N_3,10.5/N_1,11.4/N_1} {
        \node at (-0.5,\y) (R\y) {\tiny $\N$};
    }
    % column size
    \foreach \x/\N in
    {0.5/N_1,1.5/N_2,2.5/N_1,3.5/N_3,4.5/k+2,5.5/N_4,6.5/N_2,7.5/N_3,8.5/n+2,9.5/N_4,10.5/n,11.5/k} {
        \node at (\x,-0.5) (C\x) {\tiny $\N$};
    }
  \end{tikzpicture}
\end{frame}

\begin{frame}
  \frametitle{Open problems}

  \begin{block}{Combinatorics}
    How many permutations of length $2n$ are squares?

    How many $(213,231)$-avoiding
    permutations of length $2n$ are squares? (Equivalently,
    how many binary strings of length $2n$ are squares?)
  \end{block}

  \begin{block}{Combinatorics}
    Given two permutations $\pi$ and~$\sigma$, how hard is the problem of
    deciding whether $\sigma$ is a square root of~$\pi$?
  \end{block}

  \begin{block}{Algebra}
    \ldots
  \end{block}

\end{frame}
